Case based Arabic Morphological Analysis with Compositional Non-deterministic Automata

### 2010-06-26 trunk/tex/sarf.tex
9 pages

\documentclass[11pt,letterpaper]{article}
\usepackage{arabtex}
\usepackage{naaclhlt2010}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{amsthm}

@@

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance the word \RL{'a.hmadH} 
may have two valid morphological analyses. In the first
the letter \RL{'a} is a prefix and the word means ``I praise him'',
and in the second the letter \RL{'a} is part of the stem 
\RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

@@

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesely~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, 
a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
also use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance.

We hypothesize that many NLP case studies need the 
morphological analyzer to answer simple queries that do not need 
high complexity and accuracy at the low morphological analysis level 
and that can often compensate for tolerable inaccuracy at higher levels. 
For example, if the query concerns proper names and the 
prefix in question in the analysis can only connect to a verb, 
we can provide an answer without completing the analysis. 
%TODO A lso if a word such as \RL{wasiim} was name was ambiguous and we were looking for a sequence of names, then 
We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
with features demanding resolution at an abstraction level
higher than the text itself
led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our results that a 
case-based morphological analyzer that allows a case-based
controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

@@

In this paper, we present Sarf, a {\em novel case-based efficient
morphological analyzer} that uses compositional 
non-deterministic Automata driven by a case based controller.
Each alive machine in Sarf represents a valid morphological analysis. 
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
character. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one character at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
To out knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A {\em hadith} is a narration related to the prophet Mohammad
through a {\em sanad} or a sequence of narrators. 
The authentication of a hadith highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a hadith book into narrations, then segmenting each narration into
its content or {\em matn} and its sanad.
We also would like to partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books.

The hadith case study is interesting for Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text, and ignores most of the analyses 
where we do not have proper names in text as long 
as they are valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the sun of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
\item {\bf Exhaustive analysis:} Since we have a case-based 
controller that can decide on the fly whether an analysis is 
accepted or not, we can afford to keep all valid analyses modulo
those pruned by the case controller. 
We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
This allowed Sarf to retain better compositional part of speech
tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

The survey work in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzers prunes these decisions. 
Up to our knowledge, Sarf is the only morphological analyzer
that addresses the ``run-on'' words problem.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs than Buckwalter where it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix while these need to be highly 
redundant with Buckwalter and that may lead to consistency issues. 
Sarf keeps the possible analyses alive as a sequence corresponding
to the text. Buckwalter produces a set of solutions for each token 
and the several analysis threads of the whole text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules is not controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a user custom case 
controller, while with Beesley the effort requires recompiling the
Xerox rules.

@@

We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find evidence in this decision that our hypothesis
about the utility of case-based analyzers. The refinements
that resulted in algorithmic changes in SAMA were done manually
and worked in integration with the ATB format only. 
In Sarf we propose a case-based controller approach that can modify 
the behavior of the analyzer on the fly and that should interact
with the case study. In Sarf the analyzer remains general, and the 
case-based controller specializes the analyses.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
While the SVM analysis learns features from the text passed to the case
study, it does not make use of the type oh the query the 
case study targets in the text.

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries or our analysis 
to run the prefix, suffix and stem FSAs. 
We think that their inferential-realizational approach 
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can make benefit of many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammar correct.
We can compare the case-based controller in Sarf to a high level
component in the ElixirFM domain specific language.

@@

Note that $\Sigma$ can report a full word at any accept state
now by spawning a new Automata using the $\epsilon$ transition
to the prefix FSA.
This is refined in Sarf to happen only with white space, delimiters, 
or non-connecting characters. 
We left that out in Figure~\ref{f:example} for clarity.

Finally, The case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an Automata that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others.

\subsection{Recursive Affixes}
\label{sec:recaffix}

Morphological analysis partitions a given Arabic word
into a set of morphemes.
A morpheme is either an affix or a stem. 
We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\cite{Buckwalter:02} keeps separate lists 
of prefixes, suffixes and stems and assigns a category
to each morphemes. 
The relation between morphemes and categories is one 
to many.

A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The string $\alpha\beta\gamma$ partition of the string is a 
valid morphological analysis if  and only if we have
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
Moreover, the POS tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant and may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
was the category of the second morpheme in the pair.

\subsection{Affix Linear FSA}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\cite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into a FSA similar to that
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two FSAs are directed acyclic graphs. 
The absence of cycles in these FSAs reduces the 
traversal of the affix FSAs to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix FSAs encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the FSA corresponds to the last character
of a prefix in the prefix and suffix lists.
The accept states hold category state information as well
as POS and other tags.

\subsection{Stem Linear FSA}
\label{sec:stemFSA}

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
  Our understanding is that those books are historical documents with 
  public IP.}.

Our analysis accepts a book as a text stream input
and segments it into a vector of individual hadiths. 
We segment each hadith into two parts. The first part
is the sanad and it
contains the chain of narrators who related the hadith
to Mohammad. 
The second part is the matn and it constitutes the content
of the hadith. 
We are concerned in further exploring the sanad. 
In particular, we detect the chain of narrators
and the relation between each narrator, his ancestor, and 
his predecessor in the chain. 
We also detect the full name of each narrator that is
usually composed of several proper names with connectors
in between.

We built our case-based controller to target
the detection of proper names. 
This includes the task of finding compound names 
composed of several words such as \RL{`bd alr.hmn} often
appearing as ``run-on'' words.
The controller also targets words that mean narrate when
they appear in the neighborhood of multiple proper names.

\subsection{Controller}
% FSM for the controller
The FSA in Figure~\ref{f:hadith}

@@

We report our results based on a manual validation of 
a representative sample selected randomly from the text.

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
 & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
Segmentation accuracy & & 80\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time & &a & & & & & & & \\ \hline
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

\subsection{Efficiency Comparison Against Buckwalter and SAMA}

\subsection{Case Study Accuracy and Efficiency Results}

% accuracy measure:
% on names: a word is a name or not
% chains: narrator is missing from a chain
% segmentation: two hadiths are merged

>>>>

\documentclass[11pt,letterpaper]{article}
\usepackage{arabtex}
\usepackage{naaclhlt2010}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{amsthm}
\usepackage{booktabs}

@@

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance, the word \RL{'a.hmadH} 
may have two valid morphological analyses. 
The letter \RL{'a} may be prefix and the word means ``I praise him''.
The letter \RL{'a} may also be part of the stem 
\RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

@@

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesly~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, 
a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
also use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of performance.

We hypothesize that many NLP case studies need the 
morphological analyzer to answer simple queries that do not need 
high complexity and accuracy at the low morphological analysis level 
and that can often compensate at higher level for tolerable inaccuracy at low levels. 
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we do not need to further analyze the rest of the word 
to provide an answer.
We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
with features demanding resolution at an abstraction level
higher than the text itself
led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our results that a 
case-based morphological analyzer that allows a case-based
controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

@@

In this paper, we present Sarf, a {\em novel case-based efficient
morphological analyzer} that uses compositional 
non-deterministic Automata driven by a case based controller.
Each alive machine in Sarf represents a valid morphological analysis. 
Sarf keeps alive all possible analyses of the text and gives 
opportunity to the controller to intervene at every single input 
letter. 
The decisions of the case-based controller can thus prune false 
positives early in the run. 
Each alive machine in Sarf takes as input one letter at a time 
from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
Up to our knowledge, this is the first morphological analyzer that 
handles the ``run-on'' words problem.

@@

We validate our hypothesis using a case study from the Islamic 
literature. 
A {\em hadith} is a narration related to the prophet Mohammad
through a {\em sanad} or a sequence of narrators. 
The authentication of a hadith highly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a hadith book into narrations, then segmenting each narration into
its content or {\em matn} and its sanad.
We also partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books.

The hadith case study is interesting to Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text. It ignores the text where
proper names occur rarely as long 
as the text is valid.
The controller is also interested in discovering a subset of words 
that relate persons to each other such as \RL{bin} or ``the son of''
or words that mean narrate such as \RL{qaal} or ``said''. 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
and extracted from each one of them a vector of a three level deep
complex structure. 
We also showcase that the results of the accurate high level abstraction
results of the case-based analysis was not affected by a major
refinement to the analyzer that increased low level analysis accuracy.

@@

We make the following key contributions. 
\begin{enumerate}
\item {\bf Case-based analysis:}  We designed and implemented
a case-based morphological analyzer where a case-based controller
machine controls and guides the analysis. 
\item {\bf Exhaustive analysis:} Since we have a case-based 
controller that can decide on the fly whether an analysis is 
accepted or not, we can afford to keep all valid analyses modulo
those pruned by the case controller. 
We do that using non-deterministic FSAs. 
\item {\bf Recursive affixes:}
We refine the concatenative analysis of 
Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
This allows Sarf to perform better and 
maintain compositional part of speech tags.
\item {\bf Hadith case study:}
We evaluated Sarf using the hadith case study. We were able to
efficiently extract the sequence of narrators into a complex
data structure with three levels of hierarchy with high accuracy. 
\end{enumerate}

@@

The survey in~\cite{Sughaiyer:04} discusses and compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications in the morphological 
analyzer itself or use a specific set of part of speech tags
as their reference.
Sarf differs in that it is a general and rich morphological 
analyzer that reports all possible solutions. 
It is case-based in the sense that a controller that drives
the morphological analyzer prunes these decisions.

Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of affix compatibility rules to allow recursive 
affixes so that we can have prefix-prefix an suffix-suffix 
concatenations.
This allows Sarf to better maintain and propagate 
the part of speech information
tags associated with each prefix and suffix.
This information is highly 
redundant with Buckwalter and that may lead to consistency issues. 
Sarf keeps all the possible analyses alive where each analysis
corresponds to a set of decisions and positions in the text. 
Buckwalter produces a set of solutions for each token 
and the several analysis threads of the text need to be 
generated as a product of the token solutions.

Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
finite-state Automata (FSA). 
Beesley reports that the number of the Automata generated by a compiler
from Xerox rules can not be controlled and also reports a difficulty to 
compose the FSAs into a single FSA framework. 
Sarf solves the problem with a non-deterministic composition of three
FSAs that are the prefix, stem, and suffix FSAs. 
With Sarf it is easy to control the FSAs with a custom user case 
controller. Up to our understanding, Beesley  
requires recompiling the Xerox rules.

@@

We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
is the closest to our approach. SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. We find supporting evidence in this decision to our hypothesis
about the utility of a case-based analyzer. 
The refinements that resulted in algorithmic changes in SAMA were 
done manually and worked in integration with the ATB format. 
Sarf is a case-based controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with the any case study. In Sarf, the analyzer remains general, and the 
case-based controller specializes the analysis.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
While the SVM analysis learns features from the text passed to the case
study, it does not make use of the type of the query the 
case study targets in the text.

Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
of the Buckwalter analyzer. 
We also use deterministic parsing with tries in our analysis 
to run the prefix, suffix and stem FSAs. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the case-based controller
since in many cases the abstraction needed may not need the 
text to be grammatically correct.
%We can compare the case-based controller in Sarf to a high level
%component in the ElixirFM domain specific language.

@@

Note that $\Sigma$ reports a full word at any accept state
by spawning a new Automata using the $\epsilon$ transition
to the prefix FSA.
In Sarf, this only happens with white space, delimiters, 
and non-connecting letters. 
We left that out in Figure~\ref{f:example} for clarity.

Finally, the case-based controller can interfere at any point in the 
analysis to make a decision like ignoring an Automata that 
is not interesting. 
It may also decide to correct the analysis of one Automata
based on the decision of the others.

\subsection{Recursive affixes}
\label{sec:recaffix}

Morphological analysis partitions a given Arabic word
into a set of morphemes.
A morpheme is either an affix or a stem. 
We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
of prefixes, suffixes and stems and assigns a category
to each morpheme. 
The relation between morphemes and categories is one 
to many.

A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that morphemes
with category $c_1$ can be concatenated with morphemes
of category $c_2$. 
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
The $\alpha\beta\gamma$ partition of the string is a 
valid morphological analysis if  and only if we have
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \RL{wsy} is composed of three other prefixes
namely \RL{w}, \RL{s}, and \RL{y}.
The POS tag for \RL{wsy} is a concatenation
of the POS tags of each of the three morphemes. 
Keeping these morphemes separate in a list
is redundant and may produce consistency issues. 
We add two lists of compatibility rules $L_{pp}$ and
$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
respectively.
This allows us to keep a shorter list of affixes. 
We need one more additional modification to accommodate 
a category for the morphemes accepted by the $L_{pp}$
and the $L_{xx}$ rules. For that we introduce the concept
of a {\em resulting category}. 
The resulting category in all practical cases
is that of the second morpheme in the pair.

\subsection{Affix linear FSA}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and extracted our refined lists of recursive affixes.
We encoded the lists into an FSA similar to those
in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
The two FSAs are directed acyclic graphs. 
The absence of cycles in these FSAs reduces the 
traversal of the affix FSAs to a linear traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings of the word in question
and looks it up in the affix tables.

The prefix and suffix FSAs encode the $L_{pp}$ and
$L_{xx}$ compatibility lists.
An accept state in the FSA corresponds to the last letter 
of a prefix in the prefix and suffix lists.
The accept states hold category state information as well
as POS and other tags.

\subsection{Stem linear FSA}
\label{sec:stemFSA}

@@

A hadith in Islamic literature is a narration from the prophet Mohammad
related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.

Our analysis accepts a book as input
and segments it into a vector of hadiths. 
We segment each individual hadith into two parts. 
The first part is the sanad and it
contains the chain of narrators who related the hadith
from Mohammad. 
The second part is the matn or the content
of the hadith. 
We are concerned to further explore the sanad and
we detect the chain of narrators and 
the relation that links each narrator to his ancestor and 
his predecessor in the chain. 
We also detect the full name of each narrator that is
usually composed of several proper names with connectors
in between.

We built our hadith case-based controller to target
the detection of proper names. 
This includes the task of finding compound names 
composed of several words such as \RL{`bd alr.hmn} often
appearing as ``run-on'' words.
The controller also targets words that mean narrate when
they appear in the neighborhood of multiple proper names.

\subsection{Controller}
% FSM for the controller

We illustrate the hadith case-based controller FSA 
in Figure~\ref{f:hadith}. 
The transitions in the hadith FSA are excited
by inputs from the Sarf morphological analyzer. 
A NAME label means that Sarf encountered a proper name.
An NRC label means that Sarf met a narrator connector such as
\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
\RL{'a_hbar} "told" or one of their derivations. 
The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
The NISBA label corresponds to the an adjective that points to a person such 
as \RL{alma.sriy} "the Egyptian".

The symbol $\tau_{\mbox{NMC}}$ is a threshold
that corresponds to the number of tolerated name connectors 
that may occur between two names. 
The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
of name connectors collected since the controller
started looping in the state NMC\_S. 
The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
that corresponds to a relaxed tolerance measure that
the controller resorts to in case the words separating
two names were longer than $\tau_{\mbox{NMC}}$ but 
contained a name connector word such as IBN or NISBA.

The controller has four states that correspond to 
an abstract position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to the state NAME\_S whenever
NAME is reported by Sarf.
State NRC\_S means that the controller thinks it is in the context
of a narrator name after Sarf reports an NRC.
Similarly, state NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{psfigs/hadith.pstex_t}}
\caption{FSA controller of the hadith case study.}
\label{f:hadith}
}
\end{figure}

Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
before it gives up on its expectations.

We reach the NAME\_S state only when a
valid NAME is detected and we leave when no more NAME's are detected. 
The NRC\_S state can only be reached if an NRC is detected.
NMC\_S can only be reached from a NAME\_S state.

The FSA reports a valid chain of narrators when a sequence of names
connected by narrator connectors appears. 
It marks the beginning of the hadith with the beginning of the current sequence,
and marks the end of the hadith with the beginning of the next sequence. 
It also marks the chain of narrators as the sequence itself.

The definition of input labels such as NAME and IBN depends on the 
morphological analyzer. 
However, our case-based controller approach was able to perform well
under both a good analyzer and a refined one.

@@

\begin{table*}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
 & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
 & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
 Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
%Narrator accuracy & & & & & &  & & & \\ 
Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
\end{tabular}
\normalsize
\label{t:hadithresallresults}
\end{table*}

To evaluate our hypothesis of the utility of a case-based morphological analyzer,
we ran our case study in two modes, simple and refined.
In the simple mode, Sarf is controlled by the hadith controller.
In the refined mode, we apply several refinements to Sarf.
The first refinement removes female proper names since they do not occur often in
hadith.
We also add a refinement that detects compound names as they are common in narrator
names.
Finally, we consider a possessive form as a reference to a person and treat
it as a narrator.

We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
The word count row is an indicator of the size of data.
The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
narrators, and a narrator contains names.
In general, we observe that the refinements slightly improved performance and accuracy at
the lowest level of abstraction by reducing false positives.
However, the refinements had little to no effect on the high accuracy measures at higher
levels of abstraction.
We believe that this is a strong supporting evidence to our hypothesis.

The names row reports the total number of detected names. The names/narrator row reports the average
number of names per narrator structure. The narrators row reports the number of narrators detected.
The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
the number of detected chains in the book.
The ignored names row reports the number of names detected outside the chains and thus irrelevant to
the case study.
The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
The narrator name accuracy row reflects the percentage of correct proper names detected within
narrator structures.
